An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-15 EconGrowth-R-Functions.Rnw
Now including Dave Murphy's energy cost share graph. Re-organized the sections of the presentation to make more sense.

<<setup_parent, echo=FALSE, message=FALSE, eval=TRUE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
require(reshape2) # Provides access to data melting. See http://cran.r-project.org/web/packages/reshape2/reshape2.pdf
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (i.e., cache=TRUE).
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  message=FALSE, #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.
  echo=FALSE     #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
)

@@

#############################
# This function creates a series of graphs for Cobb-Douglas with energy functions
# showing various parameters as a function of gamma.
# countryAbbrev tells which country you want to view
# energyType tells which type of energy you want
# textScaling tells how you want to scale the text on the axes and other places.
##
cdeGridGraphs <- function(countryAbbrev, energyType, textScaling=1.0){
  gammaGrid <- seq(0.0, 0.95, by=0.05)
  data <- cdeGridData(countryAbbrev, energyType, gammaGrid)
  # Now melt the data so that it is in the correct form.
  data_molten <- melt(data, id="gamma")
  alphaBetaGammaRange <- c(0.0, 1.0)
  alphaBetaGammaTics <- c(0.0, 0.2, 0.4, 0.6, 0.8, 1.0)
  lambdaRange <- c(0.0, 0.004)
  lambdaTics <- c(0.0, 0.001, 0.002, 0.003, 0.004)
  sseRange <- c(0.0, 0.08)
  sseTics <- c(0.0, 0.02, 0.04, 0.06, 0.08)
  # Create a lattice graph that shows the four variables (lambda, alpha, beta, and SSE) vs. gamma.
  graph <- xyplot(value ~ gamma | variable, data=data_molten,
                  # 1=lambda, 2=alpha, 3=SSE, 4=beta
                  index.cond=list(c(3,1,2,4)), #orders the panels as SSE, lambda, alpha, beta.
                  strip = strip.custom(factor.levels=c("$\\lambda$", "$\\alpha$", "$SSE$", "$\\beta$"), 
                                       bg="transparent", # Sets background transparent to match the graph itself.
                                       par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                       ),
                  as.table=TRUE, #indexing of panels starts in upper left and goes across rows.
                  pch = c(1), col.symbol = c("black"), #Controls symbol parameters
                  scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                              tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                              x=list(relation="free",
                                     at=alphaBetaGammaTics),
                              y=list(relation="free",  #allows each axis to be different
                                     #specifies location for tics
                                     at=list(lambdaTics, alphaBetaGammaTics, sseTics, alphaBetaGammaTics))
                              ), 
                  ylim=list(lambdaRange, alphaBetaGammaRange, sseRange, alphaBetaGammaRange), #y axis limits
                  #axis labels and scaling
                  xlab=list(label="$\\gamma$", cex=textScaling), 
                  ylab=list(label="")
                  )
  return(graph)
}

>>>>

<<setup_parent, echo=FALSE, message=FALSE, eval=TRUE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
require(reshape2) # Provides access to data melting. See http://cran.r-project.org/web/packages/reshape2/reshape2.pdf
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=TRUE,
  # ********************************************************************************************************
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (i.e., cache=TRUE).
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  message=FALSE, #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.
  echo=FALSE     #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
)

@@

#############################
# This function creates a graph for Cobb-Douglas with energy functions. The graph has 
# a double Y axis. The first Y axis has alpha, beta, and gamma vs. gamma. The second Y axis
# has SSE vs. gamma.
# countryAbbrev tells which country you want to view
# energyType tells which type of energy you want
# textScaling tells how you want to scale the text on the axes and other places.
# This 
##
cdeGridGraph <- function(countryAbbrev, energyType, keyXLoc = defaultKeyXLoc, keyYLoc = defaultKeyYLoc, textScaling = 1.0){
  gammaGrid <- seq(0.0, 0.95, by=0.05)
  data <- cdeGridData(countryAbbrev, energyType, gammaGrid)
  # Now melt the data so that it is in the correct form.
  data_molten <- melt(data, id="gamma")
  alphaBetaGammaRange <- c(0.0, 1.0)
  alphaBetaGammaTics <- c(0.0, 0.2, 0.4, 0.6, 0.8, 1.0)
  lambdaRange <- c(0.0, 0.004)
  lambdaTics <- c(0.0, 0.001, 0.002, 0.003, 0.004)
  sseRange <- c(0.0, 0.08)
  sseTics <- c(0.0, 0.02, 0.04, 0.06, 0.08)
  # Create a lattice graph that shows the four variables (lambda, alpha, beta, and SSE) vs. gamma.
  graphType <- "b" #b is for both symbls and lines. Necessary so that the key looks right.
  lineTypes <- c(2, 1, 3, 0) #line types. See http://en.wikibooks.org/wiki/R_Programming/Graphics
  lineWidths <- c(1, 3, 1, 0) #line widths. 0 means no line.
  symbols <- c(NA, NA, NA, 1)
  colors <- c("black", "black", "black", "black") #line and symbol colors
  abgGraph <- xyplot(alpha+beta+gamma ~ gamma, data=data,
                     type=graphType,
                     lty = lineTypes, lwd = lineWidths, col = colors, #Controls line parameters
                     pch = symbols, col.symbol = colors, #Controls symbol parameters
                     key=list(text=list(c("$\\alpha$", "$\\beta$", "$\\gamma$", "$SSE$")),
                              type=graphType,
                              cex=keyTextSize * textScaling, #controls size of text in the key
                              lines=list(lty=lineTypes, lwd=lineWidths), #controls line types
                              pch=symbols, col=colors, #controls symbol (plot characters) types
                              columns=keyColumns, x=keyXLoc, y=keyYLoc), #controls columns and position of the key
                     scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 x=list(at=c(0.0, 0.2, 0.4, 0.6, 0.8, 1.0)), #location for x tics
                                 y=list(at=c(0.0, 0.2, 0.4, 0.6, 0.8, 1.0))  #location for y tics
                     ), 
                     ylim=c(0.0, 1.5), #y axis limits
                     #axis labels and scaling
                     xlab=list(label="$\\gamma$", cex=textScaling), 
                     ylab=list(label="$\\lambda$, $\\alpha$, $\\beta$", cex=textScaling)
  )
  sseGraph <- xyplot(SSE ~ gamma, data=data,
                     pch = c(1), col.symbol = c("black"), #Controls symbol parameters
                     scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 y=list(at=c(0.0, 0.02, 0.04, 0.06, 0.08)) #location for y tics                                          
                                 ), 
                     ylim=c(0.0, 0.08),
                     xlab=list(label="$\\gamma$", cex=textScaling), #Not used by the double-Y graph.
                     ylab=list(label="$SSE$", cex=textScaling)
                     )
  # Combine the two graphs into a double-Y graph.
  doubleYGraph <- doubleYScale(abgGraph, sseGraph, add.ylab2=TRUE, use.style=FALSE)
  return(doubleYGraph)
}

### 2013-03-15 EconGrowthSeminar.Rnw
Now including Dave Murphy's energy cost share graph. Re-organized the sections of the presentation to make more sense.

\title[Energy and Growth] % (optional, use only with long paper titles)
{An Empirical Analysis of the Role of Energy in Economic Growth}

%\subtitle
%{Presentation Subtitle} % (optional)

@@

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
% \AtBeginSubsection[]
% {
%   \begin{frame}<beamer>{Outline}
%     \tableofcontents[currentsection,currentsubsection]
%   \end{frame}
% }

@@

\begin{frame}{World Oil Supply is Tightening} %----------------------------
  \framesubtitle{\citep{Tverberg:2013}}
  \includegraphics[width=4.0in]{figure_other/Figure_Tverberg_2013.png}
\end{frame}

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{Energy is considered to be economically unimportant.}
    \item{Compare cost shares of capital stock, labor, and energy.}
  \end{itemize}
\end{frame}

\begin{frame}{Energy Cost Share and Economic Growth} %----------------------------
  \framesubtitle{\citep{Bashmakov:2007ek}}
  \begin{center}
    \includegraphics[height=6.3cm]{figure_other/Figure_Bashmakov_2007.pdf}
  \end{center}
%  \citet{Bashmakov:2007ek}
\end{frame}

\begin{frame}{Observations and Questions} %----------------------------
  \begin{itemize}
    \item{Historical economic development paths (US, Western Europe) are energy intensive.}
    \item{What happens when energy supplies are constrained and prices rise?}
    \item{Will developing countries be able to expand their economies?}
    \item{How account for energy in economic models?}
  \end{itemize}
\end{frame}

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Figure_Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{Economic output is a function of ``factors of production''}
    \item{Output defined as GDP ($Y$)}
    \item{All inputs and outputs normalized to initial year (0)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Energy ($E$)} %----------------------------
  Caleb complete
\end{frame}

\begin{frame}{Thermal Energy ($Q$)} %----------------------------
  Caleb complete
\end{frame}

\begin{frame}{Exergy ($X$)} %----------------------------
  Caleb complete
\end{frame}

\begin{frame}{Useful Work ($W$)} %----------------------------
  Caleb complete
\end{frame}

@@

\begin{frame}{Cobb-Douglas} %----------------------------
  \begin{equation*} \label{eq:CD_With_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{equation*} \label{eq:CD_Const_Returns_To_Scale}
    \alpha + \beta + \gamma = 1
  \end{equation*}
  \begin{equation*} \label{eq:CD_Constraints}
    0 \leq \alpha, \beta, \gamma \leq 1
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\lambda$   &   technological progress\\
      $\alpha$   &    ``strength'' of capital\\
      $\beta$   &    ``strength'' of labor\\
      $\gamma$   &    ``strength'' of energy\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{equation*} \label{eq:CD_With_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits data}
    \item{Almost identical to Cobb-Douglas without energy}
    \item{Perform grid search over $\gamma$}
    \item{Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint ($\alpha + \beta + \gamma = 1$)}
  \end{itemize}
\end{frame}

\begin{frame}{Grid Search Over $\gamma$}{South Africa Cobb-Douglas With Energy} %----------------------------
<<CDeGammaGraphsZA, fig.width=presentationGraph1ColWidth, fig.height=presentationGraph1RowHeight, eval=TRUE>>=
  cdeGridGraph("ZA", "Q", textScaling=0.8)
@
\end{frame}

@@

@@

\begin{frame}{Conclusions} %----------------------------

  \begin{itemize}
  \item
    Including energy is unwarranted \\
    \scriptsize
    Little statistical evidence that including energy in production functions improves fit.
    \normalsize
  \item
    Including energy is unnecessary \\
    \scriptsize
    Energy is always \emph{already} embedded in capital stock ($K$) and labor ($L$).
    \normalsize
  \vskip0pt plus.5fill
  \item
    Need a modeling method that captures energy's embeddedness \\
    \scriptsize
    System dynamics modeling
    \normalsize
  \end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Bibliography}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\title[Energy and Growth] % (optional, use only with long paper titles)
{The Role of Energy in Economic Growth}
\subtitle
{An Empirical Analysis} % (optional)

@@

% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
\AtBeginSubsection[]
{
  \begin{frame}<beamer>{Outline}
    \tableofcontents[currentsection,currentsubsection]
  \end{frame}
}

@@

\begin{frame}{World Oil Supply is Tightening}{\citep{Tverberg:2013}} %----------------------------
  \includegraphics[width=4.0in]{figure_other/Figure_Tverberg_2013.png}
\end{frame}

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \pause
    \item{Energy unimportant for classical economists}
    \vspace{\baselineskip}
    \item{Energy usually ignored}
    \pause
    \vspace{\baselineskip}
    \item{But \dots}
  \end{itemize}
\end{frame}

\begin{frame}{Energy Cost Share and Economic Growth}{\citep{Bashmakov:2007ek}} %----------------------------
  \begin{center}
    \includegraphics[height=6.3cm]{figure_other/Figure_Bashmakov_2007.pdf}
  \end{center}
\end{frame}

\begin{frame}{Energy Cost Share and Economic Growth}{\citep{Murphy:2011jh}}  %----------------------------
  \begin{center}
    \includegraphics[height=6.3cm]{figure_other/Figure_Murphy_2011.png}
  \end{center}  
\end{frame}

\begin{frame}{Observation and Questions} %----------------------------
  \begin{itemize}
    \item{Historical economic development paths (US, Western Europe) are energy intensive.}
    \vspace{\baselineskip}
    \pause
    \item{What happens when energy supplies are constrained and prices rise?}
    \item{Will developing countries be able to expand their economies?}
    \vspace{\baselineskip}
    \pause
    \item{How account for energy in economic models?}
  \end{itemize}
\end{frame}

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Figure_Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{Economic output ($Y$) is a function of ``factors of production''}
    \item{Output ($Y$) defined as GDP}
    \item{All inputs and outputs normalized to initial year (0)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Energy ($E$)} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$)}
    \item{Exergy ($X$)}
    \item{Useful work ($W$)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{No Shortage of Economic Growth Models} %----------------------------
  Single Factor: $y = \mathrm{e}^{\lambda (t-t_0)}f^{m}$
  \vspace{\baselineskip}
  \pause

  Cobb-Douglas: $y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}$
  \vspace{\baselineskip}
  \pause

  Cobb-Douglas with Energy: $y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}$
  \vspace{\baselineskip}
  \pause

  Constant Elasticity of Substitution:
  \small
  $y = \gamma \mathrm{e}^{\lambda(t-t_0)} \left\{\delta \left[\delta_1 k^{-\rho_1} + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} + (1-\delta) e^{-\rho} \right\}^{-1/\rho}$
  \normalsize
  \vspace{\baselineskip}
  \pause

  LINEX: $y = e \: \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]}$
  \vspace{\baselineskip}
  \pause

  System Dynamics Models
\end{frame}

\begin{frame}{The Question} %----------------------------
  Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Analysis]{Analysis of Growth Models}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Cobb-Douglas} %----------------------------
  \begin{equation*} \label{eq:CD_Without_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}
  \end{equation*}
  \begin{equation*} \label{eq:CD_Const_Returns_To_Scale}
    \alpha + \beta = 1
  \end{equation*}
  \begin{equation*} \label{eq:CD_Constraints}
    0 \leq \alpha, \beta \leq 1
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      \hline
      $\lambda$   &   technological progress\\
      $\alpha$   &    ``strength'' of capital\\
      $\beta$   &    ``strength'' of labor\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{Example Cobb-Douglas Fits}

Heun complete.

\begin{frame}{Cobb-Douglas With Energy} %----------------------------
  \begin{equation*} \label{eq:CDe_With_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{equation*} \label{eq:CDe_Const_Returns_To_Scale}
    \alpha + \beta + \gamma = 1
  \end{equation*}
  \begin{equation*} \label{eq:CDe_Constraints}
    0 \leq \alpha, \beta, \gamma \leq 1
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\lambda$   &   technological progress\\
      $\alpha$   &    ``strength'' of capital\\
      $\beta$   &    ``strength'' of labor\\
      $\gamma$   &    ``strength'' of energy\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{equation*}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits historical data}
    \item{Almost identical to Cobb-Douglas without energy}
    \item{Perform grid search over $\gamma$}
    \item{Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint ($\alpha + \beta + \gamma = 1$)}
  \end{itemize}
\end{frame}

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{equation*}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits data}
    \item{Almost identical to Cobb-Douglas without energy}
    \item{Perform grid search over $\gamma$}
    \item{Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint ($\alpha + \beta + \gamma = 1$)}
  \end{itemize}
\end{frame}

\begin{frame}{Grid Search Over $\gamma$}{South Africa Cobb-Douglas With Energy} %----------------------------
<<CDeGammaGraphsZA, fig.width=presentationGraph1ColWidth, fig.height=presentationGraph1RowHeight>>=
  cdeGridGraph("ZA", "Q", textScaling=0.8)
@
\end{frame}

@@

\begin{frame}{Conclusions} %----------------------------
  \begin{itemize}
  \item
    Including energy is unwarranted \\
    \scriptsize
    Little statistical evidence that including energy in production functions improves fit.
    \normalsize
  \item
    Including energy is unnecessary \\
    \scriptsize
    Energy is always \emph{already} embedded in capital stock ($K$) and labor ($L$).
    \normalsize
  \vskip0pt plus.5fill
  \item
    Need a modeling method that captures energy's embeddedness \\
    \scriptsize
    System dynamics modeling
    \normalsize
  \end{itemize}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{References}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
